### 2016-02-24 doc/draft.Rmd
sent intro to Adam

```{r message=F,echo=FALSE,cache=T,results='hide'}
xamp_roc<-"figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"

@@

knitr::knit("../../scripts/figures.Rmd",tangle = T)

```

@@

st_error="XXX"
```

@@

The popularization of next generation sequencing (NGS) is revolutionizing the study of viral diversity. Until recently, our understanding of viral evolution was limited by high-effort, low-throughput sequencing methods. Thus much of what is known regarding \emph{in vivo} viral evolution is derived from epidemiological studies (possibly cite examples, flu foot and mouth ect). However, with the advent of NGS it is now feasible to detect rare single nucleotide variants present in hundreds of patient derived samples (cite). To many, next generation sequencing promises a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite). Although NGS is highly sensitive, its specificity in the context of intrahost viral populations has not been robustly measured. Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes where by errors can be introduce and propagated. The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for in order to ensure accurate analysis.

Many sample preparation pipelines have been developed to control for errors in NGS based viral studies; hover, each approach has its own caveats and ultimately limitted application. For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej]. Rolling circle cDNA synthesis then produces on average three tandem reads of the same RNA fragment. Consensus sequences are assigned to each RNA fragment controlling for reverse transcription, PCR, and sequencing errors. However, the sensitivity of the method has not been robustly benchmarked against known rare variants. Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting viral RNA [@Acevedo:2014df]. Thus the method has limited application to patient-derived samples  that  yield low nucleic acid concentrations and are highly enriched for host RNA.

Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing erros [@Jabara:2011eo]. While this approach requires less input and targets sequencing to the viral genome it is unable to control for RT errors present in the cDNA. Additionally, primer Id based methods require that each bar code be physically attached to a PCR product, and so the region of the genome that can be bar-coded and analyzed is limited by the read length of the sequencing platform (300-500 bp for Illumina).  Therefore primer ID is useful in interrogating specific regions of a genome, but has limited application in whole genome studies. An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for the whole genome sequencing and controls for error propogated by library preparation [@Djikeng:2008cd]. RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation. This method controls for any errors that may arise in the PCR step of the library prep and controls for PCR duplicates and frequency biases. However, the bar-coding reaction used in SISPA is biased by the bar codes resulting in uneven coverage and thus sensitivity across the genome [@Rosseel:2013dn].

In addition to sample processing pipelines, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequence data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk]. In general these algorithms distinguish true variants from errors by calibrating base specific error rates across the genome. The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and the sequence context of a given base. True variants are identified as those with frequencies that exceed the expected error rate according to some predetermined statistical test. Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using viral populations, and to our knowledge none have been tested under patient-derived sample conditions. The accuracy of such algorithms in the context of NGS studies of patient-derived samples is not known.

Here, we use defined populations of influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient derived conditions. In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction prior to sequencing on the Illumina platform. PCR products are sheared using Covaris sonication which gives unbiased coverage across the genome and, variants are identified using the DeepSNV algorithm. In this work, we highlight the challenges that accompany NGS based studies of viral diversity and  make some suggestions for improving the accuracy of our approach. It is our hope that this work exemplifies the quality controls that should be run prior to any NGS based viral study and provides a comprehensive data set for benchmarking future variant calling approaches.

@@

The ability to reliably distinguish true from false single nucleotide variants (SNV) in next generation sequencing data is integral to accurate studies of viral diversity. Sensitivity is the proportion of true variants identified; while, specificity is the proportion of true negatives identified. $1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives. Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and nothing else.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but reduce sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`). An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make an ROC curve for a variant calling algorithm one must test the assay on known data. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A). These points are then used to construct the curve (`r xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

@@

A comprehensive comparison of SNV calling approaches  is beyond the scope of this work. Here, we robustly benchmark one variant caller and highlight few means for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome. Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of a sequencing error. Additionally, the control and experimental samples are processed together and have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling method, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. We created defined mixtures of two influenza strains, WSN33 and PR8, to characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus populations.

WSN33 and PR8 genomes were mixed serially such that WSN33 was present at  frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A). Because these viruses differ at 491 positions, there are 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0. Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants in the `r miseq_kb`kb of the influenza genome we analyzed.

@@

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8. Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture. To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock. We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain populations with lower input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.

The 20 mutations present in our initial viral mixture were the  only true positives considered in our analysis. Any variant that was present in both duplicates of the wild type stock was masked and  considered neither a true positive nor a true negative. This was done to avoid benchmarking our pipeline with variants identified by our pipeline.

The samples with the greatest input concentration and variant frequencies of 1\% and greater had the highest sensitivity (`r hiseq` B and C). Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples. Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was 1,000 fold less . The reduced sensitivity is likely due to the fact that we expect only 7,000 of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.

%This is a little tricky since we amplified the genomes first and then mixed them.

During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. The decreased input concentration and the accompanying increase in RT-PCR cycles is likely the cause of reduced specificity compared with that seen in the cell culture experiment. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. Increased specificity can be achieved by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

Many next generation sequencing studies utilize mapping (MapQ) and base (Phred) quality cut-offs to ensure that only the highest caliber sequencing data is used to call variants. Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct. In our initial analysis we masked bases that had a base  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut offs. We  sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs. Initially, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 but were unable to distinguish true from false positives in our $10^5$ samples (`r Qual` A). Many false positives occur on well mapped reads with high quality base calls.

We further parsed our false variant calls by locating them within individual sequencing reads. It is well known that sequence quality drops near the end of a read, and we found that our false positives clustered to these regions (`r Qual` B). The average Phred score of these false positives was xx, further demonstrating that filtering on quality score alone is insufficient. In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.

Based on these data we applied a number of empirically determined cut-offs and drastically increased our specificity without sacrificing sensitivity (`r Qual` C and D). For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read. In addition to these quality cut-offs we applied a number of other criteria in an attempt to increase accuracy. We tried using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample. All of these results have be summarized in an interactive shiny application available for download at $https://github.com/lauringlab/benchmarking_shiny.git$  We invite the reader to explore the affects of all these criteria on our data set.

>>>>

```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
xamp_roc<-"figure 1"
Xamp_roc<-"Figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"

@@

knitr::knit("../scripts/figures.Rmd",tangle = T)
source("./figures.R")
#this is so the cache resets
```

@@

st_error="XXX"
false_p_phred<-round(false_p_phred,digits=1)
#csl: journal-of-virology.csl
```

@@

Until recently, our understanding of intrahost viral evolution was limited by high-effort, low-throughput sequencing methods. Much of what is known regarding \emph{in vivo} viral evolution has been derived from epidemiological studies (possibly cite examples, flu foot and mouth ect). However, with the advent of next generation sequencing (NGS) it is now feasible to detect rare single nucleotide variants present in hundreds of patient-derived samples (cite examples?). Next generation sequencing offers a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite examples?). Although NGS is highly sensitive, its specificity in the context of intrahost viral populations has not been robustly measured. Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduce and propagated. The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.

Many sample preparation pipelines have been developed to control for errors in NGS based studies of RNA virus populations; however, each approach has its own caveats that ultimately limit its application. For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej]. Rolling circle cDNA synthesis then produces tandem reads of the same RNA fragment. Consensus sequences are assigned to each RNA fragment controlling for reverse transcription, PCR, and sequencing errors. However, the sensitivity of the method has not been robustly benchmarked against known rare variants. Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting sequencing to viral RNA [@Acevedo:2014df]. As such, the method has limited application to patient-derived samples that are often characterized by low nucleic acid concentrations and are enriched for host RNA.

Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing errors [@Jabara:2011eo]. This approach requires less input and targets sequencing to the viral genome, but unlike Cirseq, it is unable to account for RT errors present in the cDNA. Additionally, primer Id based methods require each bar code be physically attached to a PCR product, and so the region of the genome that can be targeted is limited by the read length of the sequencing platform (250-500 bp for Illumina).  Primer ID based methods are useful in interrogating specific regions of a genome (cite examples?), but have limited application in whole genome sequencing. An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd]. RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation. SISPA controls for any errors that may arise during the library amplification and controls for PCR duplicate frequency biases. However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome [@Rosseel:2013dn].

Given the limitations described above, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequencing data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk]. In general these algorithms calibrate base specific error rates across the genome. The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and sequence context. True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test. Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under patient-derived sample conditions. The accuracy of such algorithms in the context of NGS studies of patient-derived RNA viral populations is not known.

We are interested in using NGS to study the intrahost dynamics of RNA viruses. Because our experimental design relies on whole genome sequencing of patient-derived samples, we are unable to use any of the above error-controlling sample preparation pipelines. We, are left with only statistical algorithms to control for errors. However, the accuracy of these algorithms under patient-derived sample conditions is unknown. Here, we use defined populations of Influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient-derived conditions. In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction, sheared using Covaris sonication, sequenced on the Illumina platform, and analyzed using the DeepSNV algorithm and in house scripts (all available at \emph{https://github.com/lauringlab/variant_pipeline})[@Gerstung:1js]. In this work, we highlight the challenges that accompany NGS based studies of viral diversity and  make some suggestions for improving the accuracy of our approach. This work exemplifies the quality controls that should be run prior to any NGS based study of RNA virus populations and provides a comprehensive data set for benchmarking future variant calling approaches.

@@

The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity. The accuracy of any SNV calling pipeline can be described in terms of sensitivity and specificity. Sensitivity is the proportion of true variants properly identified; while, specificity is the proportion of true negatives properly identified. Specificity is often reported as 1-specificity or the false positive rate. The false positive rate gives the ratio of false positive variants to true negatives. Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but often reduces sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r Xamp_roc`). An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make an ROC curve for a variant calling pipeline one must test the assay on known data. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc` A). These points are then used to construct the curve (`r Xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

@@

A comprehensive comparison of SNV calling approaches is beyond the scope of this work. Here, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome. Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of an error. Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV when applied to RT-PCR amplified RNA virus populations.

WSN33 and PR8 were plaque purified and sanger sequenced. cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A). These viruses differ at 491 positions (primer sites were excluded from analysis), providing 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0. Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb`kb of the influenza genome we analyzed (3 variants per genomic position).

@@

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8 (cite). Also the nucleic acid concentration in patient samples is much lower than that found in samples grown in cell culture (data not shown). To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock. Accross all 20 clones 4 additional mutations were found (frequencies 1.2%-3.7%). We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.

The 20 mutations present in our initial viral mixture were the  only true positives considered in our analysis. Any variant that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative. This was done to avoid benchmarking our pipeline with variants identified by our pipeline.

The samples with the greatest input concentration and variant frequencies of 1\% and greater had the highest sensitivity (`r hiseq` B and C). Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples. Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.  The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above. During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome (again primer sites were excluded). Increased specificity can be achieved by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

Many next generation sequencing studies utilize mapping (MapQ) and base (Phred) quality cut-offs to ensure that only the highest caliber sequencing data is used to call variants. Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct. In our initial analysis we masked bases that had a Phred  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut-offs. We  sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs. Initially, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 but were unable to distinguish true from false positives in our $10^5$ samples (`r Qual` A). Many false positives occur on well mapped reads with high quality base calls.

We further parsed our false variant calls by locating them within individual sequencing reads. It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered to these regions (`r Qual` B). The average Phred score of these false positives was `r false_p_phred`, further demonstrating that filtering on quality score alone is insufficient. In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.

Based on these data we applied a number of empirically determined cut-offs and drastically increased our specificity without sacrificing sensitivity (`r Qual` C and D). For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read. In addition to these quality cut-offs we applied a number of other criteria in an attempt to increase accuracy. We tried using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample. All of these results have be summarized in an interactive shiny application available for download at \emph{https://github.com/lauringlab/benchmarking_shiny.git}  We invite the reader to explore the affects of all these criteria on our data set.
